Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-09 manuscript/main.tex
yingying matrix notation

	\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}
In practice, the functional predictors $X^{(1)}, \ldots, X^{(K)}$ are not observed as complete curves, but only through their values at a finite set of time points, which may differ across the $K$ predictors. Thus, each observation is represented by a finite-dimensional vector $X_i^{(k)}(\mathbf{t}_k) \in \mathbb{R}^{|\mathbf{t}_k|}$, where $\mathbf{t}_k \subset \mathcal{T}_k$. The first step in functional data analysis typically involves reconstructing the underlying functional form from these discrete measurements, commonly by assuming that sample paths lie in a finite-dimensional space spanned by basis functions.

Given an orthonormal basis $\{b_l(t)\}$, the functional predictors and the corresponding regression coefficients can be decomposed into
\begin{align*}
	X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t), \quad 
	\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t),
\end{align*}
where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively. While it is possible to use a different basis for each functional predictor, we adopt a common basis system for all predictors for notational simplicity.

For $k = 1, \ldots, K$, let $\Phi^{(k)}   := \{\phi^{(k)}_1, \ldots, \phi^{(k)}_M\}$, with $M \geq 1$, be an $M$-dimensional basis in $L^2(\mathcal{T}_k)$. In our proposed hybrid PLS regression algorithm, this basis system is used to approximate both the predictors and the PLS components. Additionally, it serves as a regularization mechanism, borrowing strength across components and helping to prevent overfitting. 	The number of basis functions $M$ for each functional predictor can be set to a sufficiently large value (e.g., 15 or 20) to incorporate a rich variety of patterns that functional predictors may exhibit. Note that there is no need to carefully choose different $M$ values for different functional predictors, as an appropriate penalization scheme will be introduced in the later section to regularize each functional predictor to achieve a desired smoothness (see Section \ref{sec: Regularization of PLS Components}).

For $k = 1, \ldots, K$, we project the $k$th functional predictor $X_i^{(k)}$ onto 
$\operatorname{span}(\Phi^{(k)})$:
\begin{equation*}
	X_i^{(k)} \approx \sum_{m=1}^M c_{im}^{(k)} \phi_m^{(k)}, \quad i = 1, \ldots, n,
\end{equation*}
where the coefficients $c_{im}^{(k)}$ satisfy
\begin{equation*}
	X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} \phi_m^{(k)}(t_k), \quad \text{for all } t_k \in \mathbf{t}_k \subset \mathcal{T}_k.
\end{equation*}
Thus, the $n$ observations of the $k$th functional predictor can be represented by an $n \times M$ matrix of coefficients, 	where rows index samples and columns index basis functions:
\begin{equation*}
		\widetilde{\mathbf{C}}^{(k)} := ( c_{im}^{(k)} ).
\end{equation*}

Importantly, the core computation of our algorithm involves the inner product between $n$ hybrid predictor observations and a single hybrid object (the PLS direction), defined via the inner product in Definition \ref{def:hilbert_space}.
 With the basis expansion, this reduces to simple matrix operations. 
For each $k = 1, \ldots, K$, let $\boldsymbol{\Phi}^{(k)} := (\phi^{(k)}_{m \ell })$
denote the	$M \times M$ symmetric matrix with entries  the inner products between basis functions 
\begin{equation}\label{def:gram_basis}
\phi^{(k)}_{m \ell } 
:= 
\langle \phi^{(k)}_m,  \phi^{(k)}_\ell \rangle
=  
\int_{\mathcal{T}_k} \phi^{(k)}_m(t_k) \, \phi^{(k)}_\ell(t_k) \, dt_k.
\end{equation}

>>>>

	\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}
 Thus, each observation is represented by a finite-dimensional vector $X_i^{(k)}(\mathbf{t}_k) \in \mathbb{R}^{|\mathbf{t}_k|}$, where $\mathbf{t}_k \subset \mathcal{T}_k$.

Given an orthonormal basis $\{b_l(t)\}$ of $L^2([0,1])$, the functional predictors and the corresponding regression coefficients can be decomposed into
\begin{equation*}
	X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t), \quad 
	\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t),
\end{equation*}
where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively.
In practice, functional predictors are observed only at a finite set of time points rather than as complete curves. Rather than reconstructing infinite-dimensional object stated above via finite observations, we approximate each predictor using a finite-dimensional projection by truncating its basis expansion at a finite number $M$. This $M$ can be set to a sufficiently large value (e.g., 15 or 20) to capture a wide range of functional patterns. 
Notably, it is unnecessary to fine-tune $M$, as a  penalization scheme introduced in Section \ref{sec: Regularization of PLS Components} regularizes each functional component to achieve the desired smoothness. We denote the projected predictors and coefficients as
\begin{equation*}
	\tilde{X}_{ij}(t) = \sum_{l=1}^M \theta_{ijl} \, b_l(t), \quad 
	\tilde{\beta}_j(t) = \sum_{l=1}^M \eta_{0,jl} \, b_l(t).
\end{equation*}
Write $\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top$ and  $\boldsymbol{\eta}_{0j} := (\eta_{0,j1}, \ldots, \eta_{0,jM})^\top$. 
Let $\Theta_j$ be an $n \times M$ matrix whose rows are given by $\{\boldsymbol{\theta}_{ij}, \; i = 1, \ldots, n\}$.

\begin{remark}
 While it is possible to use a different basis for each functional predictor, we adopt a common basis system for all predictors for notational simplicity.
\end{remark}

 Additionally, it serves as a regularization mechanism, borrowing strength across components and helping to prevent overfitting. 	T
